---
title: "Lab #2: Ames Housing Data"
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit the README.Rmd file -->


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


**Our analysis of the Ames housing data revealed these findings:**

Sale prices can range from 0 to over 4.5 million with most properties being between $100,000-$500,000.

There are some outliers with properties that have $0 sale price which are most likely errors in the dataset.

Property characteristics like living area, bedrooms, and neighborhood strongly affect sale prices.

Newer properties and those with air conditioning have higher pricing on average.

The distribution of price differs from neighborhood to neighborhood, with some areas having median prices 3x higher than others


**Step 1 - Variables and What They Mean** 

```{r, echo=FALSE}
library(classdata)
library(dplyr)
library(ggplot2)

#inspecting data
data(ames)
head(ames)
str(ames)
```

**Here are the 16 variables and what they mean in this dataset:**

Parcel ID (Character): A unique identifier for each property parcel in the form of a 10-digit code. 

Address (Character): The street address of each property including city. 

Style (Factor): A category description with 12 types that describe the style of the property.

Occupancy (Factor): A category description with 5 types that describes the type of residential housing it is.

Sale Date (Date): The date when the property sale happened, formatted as year-month-day.

Sale Price (Numeric): The price of the property in US dollars and it ranges anywhere from $0 to $4,496,640 (of course the $0 entries are outlines/anomalies).

Multi Sale (Character): A variable for multiple sales. This seems to appears to have mostly "NA" entries.

Year Built (Numeric): The year the property was built. This ranges from the 1800s to present. 

Acres (Numeric): The lot size measured in acre and this ranges from small values like (0.023) to larger properties.

Total Living Area (sf) (Numeric): The actual living space in square feet. Ranges quite largerly from property to property.

Bedrooms (Numeric): The number of bedrooms in each property. This typically ranges from 1 to 5+ bedrooms.

Finished Bsmt Area (sf) (Numeric): The finished basement area in square feet and the many "NA" values means there might be no basement.

Lot Area (sf) (Numeric): The total lot area in square feet, and seems to correlate strongly with acres.

AC (Character): Air conditioning variable with "Yes" or "No" values to describe if the property has AC or not

FirePlace (Character): A variable with "Yes" or "No" values to describe if fire place is present in property.

Neighborhood (Factor): Geographic categorical type variable with 42 different neighborhood codes representing areas within Ames.

In conclusion, this dataset has 6,935 entries that are made up of house related measurements, categories, and character descriptors that all work to describe the property characteristics and sales associated with them in Ames since 2017.

**Step 2 - Variable Of Interest** 
Yes, Sale Price is the variable of special interest/focus
```{r, echo=FALSE}
#Summary of the statistics for Sale Price
summary(ames$`Sale Price`)
```


**Step 3 - Histogram Showing Sale Price** 

```{r, echo=FALSE}
# updated histogram to better see the distribution
ggplot(ames %>% filter(`Sale Price` > 0), aes(x = `Sale Price`)) +
  geom_histogram(bins = 50, fill = "red", color = "blue", alpha = 0.8) +
  scale_x_log10(labels = scales::dollar) +
  labs(title = "Distribution of Sale Prices",
       x = "Sale Price ($)",
       y = "Count") +
  theme_minimal()

```

#
**Step 4 - Group Members Work**

**Ivy's Work:**
```{r}
ggplot(ames, aes(x = `Style`, y = `Sale Price`)) +
  geom_boxplot() +
  coord_flip()
```

The original boxplot of Sale Price by Style showed that there were a few extreme outliers, especially in the 1 Story Frame category. These outliers, with sale prices well above typical values, caused the rest of the data to appear squashed near the bottom, making the plot difficult to interpret.

To address this, I filtered the dataset to only include houses with Sale Price below $1,000,000. This removed the extreme values and produced a clearer, more readable boxplot. With the adjustment, the differences in median sale prices across styles became more apparent, while still capturing the overall variability within each category.

```{r}
ggplot(ames %>% filter(`Sale Price` < 1000000), aes(x = reorder(`Style`, `Sale Price`, median),y = `Sale Price`))+
  geom_boxplot() +
  coord_flip()
```

The adjusted plot looks much clearer after removing extreme outliers above $1,000,000. However, I also noticed that the distribution included houses with a Sale Price of 0, which is not realistic in real life and likely represents data entry errors or missing values.

To improve data quality, I filtered out these observations by keeping only houses with Sale Price greater than 0. This results in a more meaningful visualization, since all remaining values correspond to valid housing transactions.


```{r}
ggplot(ames %>% filter(`Sale Price` < 1000000 & `Sale Price` > 0), aes(x = reorder(`Style`, `Sale Price`, median),y = `Sale Price`))+
  geom_boxplot() +
  coord_flip()
```

And this will be the final plot. 

#Interpretation:

The dataset includes many different housing styles (e.g., 2 Story Brick, 2 Story Frame, 1 Story Frame, Split Foyer Frame). In the final plot, the categories are reordered by their median Sale Price, making it easier to compare which styles tend to be more expensive.

#Overall pattern:

2 Story Brick homes have the highest median sale prices, standing out from the other styles.

Styles such as Split Foyer Frame and 1 3/4 Story Frame generally sell for lower prices.

The NA category displays unusually high variation and extreme values, likely due to data entry errors or missing style information.

Within each style, there is a substantial spread in sale prices, reflected in the box lengths and outliers, which suggests that factors beyond style (such as living area, year built and etc.) also strongly influence housing prices.

**Owen's work:**

```{r}
# Owen Arnold

range(ames$Bedrooms, na.rm = TRUE)
ggplot(ames, aes(x = Bedrooms)) + geom_histogram(bins = 30, na.rm = TRUE, fill = "red") + ggtitle("Range of Number of Bedrooms")

ggplot(ames[ames$Bedrooms > 6, ], aes(x = Bedrooms)) + geom_histogram(na.rm = TRUE, fill = "red", bins = 4) + ggtitle("Range of Number of Bedrooms")

ggplot(ames %>% filter(`Sale Price` > 0 ), aes(x = factor(Bedrooms), y = `Sale Price`)) + geom_boxplot(na.rm = TRUE, outlier.color = "red") + scale_y_log10()

```



The range of number of bedrooms is 1 - 10, most are within the range of 2 - 5.

The number of bedrooms does not have much correlation to the sale price. There are many outliers in the sale price with some houses being far more expensive than others regardless of the number of bedrooms.



**Lexi's Work:**

## what is the range of that variable? plot. describe the pattern.

```{r}

# Histogram by Decade
ames %>%
  filter(!is.na(YearBuilt), YearBuilt > 1800) %>%
  mutate(decade = (YearBuilt %/% 10) * 10) %>%
  ggplot(aes(x = factor(decade))) +
  geom_bar(fill = "blue", color = "blue", alpha = 0.7) +
  labs(title = "Number of Houses Built by Decade",
       x = "YearBuilt (by decade)",
       y = "Count of Houses") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

### Interpretation:
The variable YearBuilt ranges roughly from the late 1800s to 2010 (with some missing values). The histogram shows construction booms around the 1920s–1930s and again in the 2000s, while other decades had fewer houses built.


```{r}


# 1) Clean scatter with a smoother (good default)
ames_clean <- ames |> 
  dplyr::filter(!is.na(YearBuilt), YearBuilt > 1900, `Sale Price` > 0)

ggplot(ames_clean, aes(x = YearBuilt, y = `Sale Price`)) +
  geom_point(alpha = 0.25, size = 1) +
  geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs"), se = TRUE) +
  scale_y_log10(labels = scales::dollar) +
  scale_x_continuous(limits = c(1900, NA)) +
  labs(title = "Sale Price vs. Year Built",
       x = "Year Built",
       y = "Sale Price (log scale)") +
  theme_minimal()




```

### Interpretation:
The scatterplot shows the relationship between YearBuilt and the main variable, Sale Price. The overall pattern suggests that newer houses tend to sell for higher prices. Houses built before about 1950 cluster at lower prices, while those built between 1975 and 2005 show a clear increase in sale price. After 2010, the trend levels off or dips slightly, which may reflect fewer recent transactions in the dataset.

This variable also helps explain some of the oddities discovered in Step 3. For example, there are older homes (pre-1900) with unusually high sale prices. These may represent historic properties or extensively renovated houses that deviate from the general trend. Additionally, there are some recently built homes with very low or zero sale prices, which could be recording errors, distressed sales, or non-arm’s length transactions.
Overall, YearBuilt has a strong and intuitive relationship with sale price: newer homes generally command higher values, but the presence of both high-priced old homes and low-priced new homes shows that other factors (location, renovations, condition) also play a role.



**Marcus Work**
variable = TotalLivingArea


what is the range of that variable? plot. describe the pattern.

```{r}
rng <- range(ames$`TotalLivingArea (sf)`, na.rm = TRUE)
rng
diff(rng)
```



The range is (0:6007) square feet


what is the relationship to the main variable?
  
```{r}
plot_df <- ames |>
  filter("Sale Price" > 100, !is.na(`Sale Price`), !is.na(`TotalLivingArea (sf)`))

r <- cor(plot_df$`Sale Price`, plot_df$`TotalLivingArea (sf)`, use = "complete.obs")
```


```{r}
#full plot
ggplot(plot_df, aes(x = `TotalLivingArea (sf)`, y = `Sale Price`)) +
  geom_point(alpha = 0.25, size = 1.2) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_y_continuous(labels = scales::label_dollar()) +
  coord_cartesian(ylim = c(0, 2e6)) +
  labs(title = "Sale Price vs. Total Living Area",
       x = "Total Living Area (sf)", y = "Sale Price ($)",
       caption = paste0("Pearson r = ", round(r, 3)))
```


```{r}
#zoomed in
ggplot(plot_df, aes(x = `TotalLivingArea (sf)`, y = `Sale Price`)) +
  geom_point(alpha = 0.25, size = 1.2) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_y_continuous(labels = scales::label_dollar()) +
  coord_cartesian(ylim = c(0, 2e6)) +
  labs(
    title = "Sale Price vs. Total Living Area",
    x = "Total Living Area",
    y = "Sale Price",
    caption = paste0("Pearson r = ", round(r, 3))
  )
```

There is a small correlation between TotalLivingArea and Sale Price with an r of -0.184. as shown 
in the plot. There are about a dozen outliers with a 13,000,000+ sale price and less than 1500 
TotalLivingArea which isn't consistent with the rest of the data. you can see these in the top graph, but it is impossible to get much information because it isnt zoomed in. Theres also a good ammount of 0 sf areas in the dataset. so overall, theres a small correlation statistically, but with the zoom in plot you can see that it is positivley correlated with the vast majority of the dataset but those outliers heavily effect the r value.

**Anthony's Work**

```{r}
# Anthony's analysis: Occupancy vs Sale Price

# Check range and levels
table(ames$Occupancy)

# Distribution of occupancy types
ggplot(ames, aes(x = Occupancy)) +
  geom_bar(fill = "orange", alpha = 0.8) +
  labs(title = "Distribution of Occupancy Types",
       x = "Occupancy Type", y = "Count") +
  theme_minimal() +
  coord_flip()

# Relationship with Sale Price (exclude 0 prices)
ggplot(ames %>% filter(`Sale Price` > 0),
       aes(x = Occupancy, y = `Sale Price`)) +
  geom_boxplot(fill = "blue", alpha = 0.7) +
  scale_y_log10(labels = scales::dollar) +
  labs(title = "Sale Price by Occupancy Type",
       x = "Occupancy", y = "Sale Price (log scale)") +
  theme_minimal() +
  coord_flip()
```

**Interpretation**

Occupancy has about 5 categories (Single Family, Condo, Townhouse, Duplex, Apartment). The dataset is mostly made up of single family homes, while other categories make up a much smaller part.

The bar chart confirms that single family homes are the most common, with condos and apartments far less frequent.

**Relationship to Sale Price:**

Single family homes show the largest spread of sale prices and the highest median values.

Condos and townhouses cluster lower, with little high priced outliers.

Duplexes and apartments usually sell at lower prices compared to other types.

Some condos and apartments show really high or low prices, potentially due to bad data collection/errors.

Occupancy type does affect sale price. Single family homes make up most of the market and have the widest range of prices, while other housing types are less common and usually sell for lower/more consistent prices
